---
layout: post
title: "Download Shapefiles - Take 2"
tags: ['Shapefiles','httr']
categories: ['R']
---

```{r setup, include=FALSE, echo=FALSE}
library("miscPackage")
library("RCurl")
library("httr")
library("sp")
library("rgdal")
```

So back in 2013 I [posted a little function](https://landeco2point0.wordpress.com/2013/09/30/an-r-function-to-download-shapefiles/) I wrote for grabbing all the relevant files that make up a shapefile from a URL.  Turns out it doesn't play so well with Windows 7 or Windows 8 (HT: John Lewis).  Below is a reprised version that at least works on Ubuntu 14.04 and Windows 7.  Haven't tested it beyond that and supressing the warnings to get `httr::GET` to not complain too much about FTP seems a bit unclean.  Well, you get what you pay for. 

For all this to run you'll need `RCurl`, `httr`, `sp`, and `rgdal`.

```{r download_shp, eval=FALSE}
download_shp<-function(shape_url,layer,outfolder=".")
{
  #written by: jw hollister
  #Oct 10, 2012
  
  #updated: June 15, 2015
  
  #set-up/clean-up variables
  if(length(grep("/$",shape_url))==0)
  {
    shape_url<-paste(shape_url,"/",sep="")
  }
  #creates vector of all possible shapefile extensions
  shapefile_ext<-c(".shp",".shx",".dbf",".prj",".sbn",".sbx",
                   ".shp.xml",".fbn",".fbx",".ain",".aih",".ixs",
                   ".mxs",".atx",".cpg")
  
  #Check which shapefile files exist
  if(require(RCurl))
  {
    xurl<-getURL(shape_url)
    xlogic<-NULL
    for(i in paste(layer,shapefile_ext,sep=""))
    {
      xlogic<-c(xlogic,grepl(i,xurl))
    }
    
    #Set-up list of shapefiles to download
    shapefiles<-paste(shape_url,layer,shapefile_ext,sep="")[xlogic]
    #Set-up output file names
    outfiles<-paste(outfolder,"/",layer,shapefile_ext,sep="")[xlogic]   
  }
  #Download all shapefiles
  if(sum(xlogic)>0)
  {
    for(i in 1:length(shapefiles))
    {
      x<-suppressWarnings(httr::GET(shapefiles[i],
                                    httr::write_disk(outfiles[i],
                                                     overwrite = TRUE)))
      dwnld_file <- strsplit(shapefiles[i],"/")[[1]]
      dwnld_file <- dwnld_file[length(dwnld_file)]
      print(paste0("Downloaded ", dwnld_file, " to ", outfiles[i],"."))
    }
  } else
  {
    stop("An Error has occured with the input URL
         or name of shapefile")
  }
  }

```

And to see that it works again:

```{r run_it}
#Download the NH State Boundaries
download_shp("ftp://ftp.granit.sr.unh.edu/pub/GRANIT_Data/Vector_Data/Administrative_and_Political_Boundaries/d-nhsenatedists/2012",
                   "NHSenateDists2012")
#Read shapefiles in SpatialPolygonsDataFrame
NHBnd<-readOGR(".","NHSenateDists2012")
#Plot it
plot(NHBnd)
```
